---
title: "Drug-drug interaction screening with Omega"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drug-drug interaction screening with Omega}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pvOmega)
```

## What Omega measures

Omega (Noren et al. 2008) is a disproportionality measure for
drug-drug-event triplets. It compares the observed number of reports
listing two drugs and an adverse event with the number expected if the two
drugs did not interact.

"No interaction" is defined by an **additive** risk model: the risk of the
event under both drugs is the background risk plus the risk attributable to
each drug. This is the key difference from logistic regression, whose
interaction term assumes multiplicative risks and misses well-established
interactions such as cerivastatin with gemfibrozil.

The steps are:

1. Count the reports in each of the four exposure strata: neither drug, D1
   only, D2 only, both drugs.
2. Compute the relative reporting rate of the event in each stratum:
   `f00`, `f10`, `f01`, `f11`.
3. Estimate the expected rate under both drugs, `g11`, from `f00`, `f10` and
   `f01` only (eq. 16 of the paper).
4. Compute the expected count `E111 = g11 * n11.`.
5. Compute the shrunk ratio `omega = log2((n111 + 0.5) / (E111 + 0.5))` and
   its Gamma credibility interval.

A triplet is highlighted when `omega_lower` (Omega025) is above 0. With the
default shrinkage, at least three reports are needed for this.

## Starting from counts

`omega_from_counts()` works with any data source. The example below is the
first case study of the paper: delayed bleeding with itraconazole and oral
contraceptives.

```{r}
res <- omega_from_counts(
  n111 = 10, n11. = 23, n1.1 = 10, n.11 = 19,
  n1.. = 39, n.1. = 1489, n..1 = 39, n... = 5503
)
res[, .(f00, f10, f01, f11, g11, E111, omega, omega_lower, omega_upper)]
```


## Starting from DiAna FAERS data

With the DiAna data imported, `omega_analysis()` computes all counts and
returns one row per drug1-drug2-event combination.

```{r, eval = FALSE}
library(DiAna)
FAERS_version <- "24Q4"
import("DEMO")
import("DRUG")
import("REAC")

res <- omega_analysis(
  drug1_selected = "miconazole",
  drug2_selected = "warfarin",
  reac_selected = c("international normalised ratio increased", "haemorrhage"),
  restriction = Demo[!RB_duplicates_only_susp]$primaryid
)
res[, .(drug1, drug2, event, f11, g11, label_omega, omega_signal)]
```

Groups of terms can be collapsed with lists, as in DiAna:

```{r, eval = FALSE}
omega_analysis(
  drug1_selected = list(statins = c("simvastatin", "atorvastatin")),
  drug2_selected = "clarithromycin",
  reac_selected = list(muscle = c("rhabdomyolysis", "myopathy", "myalgia"))
)
```

## Methodological choices to report

Three choices change the counts, and should be stated in any publication:

* **Drug roles** (`drug_roles`). By default only primary suspect, secondary
  suspect and interacting drugs count as exposure, as at Uppsala Monitoring
  Centre. Reports where a drug is only concomitant fall into the background
  stratum.
* **Deduplication** (`restriction`). FAERS contains duplicate reports. Pass
  the primary IDs retained by one of DiAna's deduplication algorithms.
* **Report universe**. The total `n...` is the number of distinct reports in
  `temp_reac` (within `restriction`). Every count refers to this population.

## Interpreting results

As recommended in the original paper, do not read Omega in isolation. The
columns `f00`, `f10`, `f01` and `f11` show why a combination was
highlighted. For example, a large `f10` means D1 alone already explains much
of the reporting.

The option `no_background = TRUE` adds `omega_nb`, which assumes no
background risk. A large drop from `omega` to `omega_nb` means the signal
depends on the estimated background rate.

Omega cannot be estimated when one of the four D1/D2 strata is empty (for
example, one drug is never reported without the other), or when every report
in the D1-only, D2-only or neither-drug stratum lists the event (`f10`, `f01`
or `f00` equal to 1). Such rows return `NA` for `g11`, `E111` and the Omega
columns, with a warning that names the affected rows and the reason.


## References

Noren GN, Sundberg R, Bate A, Edwards IR. A statistical methodology for
drug-drug interaction surveillance. *Stat Med*. 2008;27(16):3057-70.

Hult S, Sartori D, Bergvall T, et al. A feasibility study of drug-drug
interaction signal detection in regular pharmacovigilance. *Drug Saf*.
2020;43:775-85.
